 

### 2021-07-21 vertical-thermal-review.Rmd
small spelling edits

Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased $VPD$ causing stomatal closure [@lloydEffectsRisingTemperatures2008]. Photosynthesis has a peaked response to temperature, with the peak commonly found to correspond to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017]. Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g. @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020 [temperate/boreal refs needed), and eventually due to biochemical constraints (@kumarathungeAcclimationAdaptationComponents2019a; Sage & Kubien 2007; Varhammer et al. 2015). rreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
[Baker 2008; @feeley_thermal_2020].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
NSC reserves also increase with tree size, such that larger trees have more buffer against biological and biophyical stressors [@niinemetsResponsesForestTrees2010a].
As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021].

@@

There are, however, noteable exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory].
Similarly, in tropical forests higher night-time temperatures have been associated with decelerating tree growth rates predominantly in smaller trees *but also in taller trees* [*Clark 2010; 2013; Feeley et al. 2007*].
Potential mechanisms for these patterns include increases in nighttime respiratory carbon loss that disproportionately reduced carbon availability for growth among small trees, or that understory $g_s$  might be stifled under humid conditions, resulting in higher $T_{leaf}$ (Fig. `r fig_leaf_T`) and lower photosynthesis. 
It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.

>>>>

Temperature can affect photosynthesis via direct and indirect pathways--directly, by altering photosynthetic enzyme activity and electron transport chain, and indirectly through increased $VPD$ causing stomatal closure [@lloydEffectsRisingTemperatures2008]. Photosynthesis has a peaked response to temperature, with the peak commonly found to correspond to the prevalent ambient growing season conditions [@tanOptimumAirTemperature2017; @slotSituTemperatureResponse2017]. Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g. @slotSituTemperatureResponse2017; @smithEmpiricalEvidenceResilience2020 [temperate/boreal refs needed), and eventually due to biochemical constraints [@kumarathungeAcclimationAdaptationComponents2019a; @sageTemperatureResponseC32007; @varhammarPhotosyntheticTemperatureResponses2015]. rreversible damage to photosystem II, leading to leaf necrosis and eventually leaf death occurs at $T_{leaf}$ of $40-60^{\circ}C$
[Baker 2008; @feeley_thermal_2020].

@@

**Vertical gradients affect tree metabolism and growth.**
Whole-tree transpiration and gross primary production (GPP) increase in tandem with increasing tree size, where size is most commonly measured diameter breast height (DBH), but functionally linked to metrics such as foliage biomass, crown volume, or height, all of which scale with DBH [**REFS**, @anderson-teixeiraSizerelatedScalingTree2015].
Specifically, increases are linked increased leaf area and in part to increasing probability that the crown is in the canopy [@muller-landauTestingMetabolicEcology2006], where higher light availability and associated elevation of $g_s$ and $A_A$ result in higher leaf area-specific transpiration and photosynthesis (Table 2).
The net foliar photosynthate production is allocated among functions including respiration, woody growth, foliar turnover, belowground allocation, reproduction, defense, and storage of non-structural carbohydrates (NSCs). 
The proportion of GPP allocated to many of these increases with tree size, including ... foliar respiration (Table 2), reproduction [**REFS**; @westIncreasingRespiratoryCosts2020].
NSC reserves also increase with tree size, such that larger trees have more buffer against biological and biophyical stressors [@niinemetsResponsesForestTrees2010a].
As a result of increasing C allocation to these other functions, mass-specific -- and sometimes even absolute -- woody growth rates decrease with tree size [**REFS**; @anderson-teixeiraJointEffectsClimate2021; @anderson-teixeiraCarbonCyclingMature2021].

@@

There are, however, noteable exceptions to the pattern of greater growth reductions under high $T_{air}$ among larger trees.
In eight mesic forests across the northeast United States, tree-ring records showed steep growth declines of understory trees at higher $T_{air}$, contrasting with the responses of canopy trees [Fig `r fig_treerings`, @rollinsonClimateSensitivityUnderstory].
Similarly, in tropical forests higher night-time temperatures have been associated with decelerating tree growth rates predominantly in smaller trees *but also in taller trees* [@clarkTropicalRainForest2003; @clarkAnnualWoodProduction2010; @feeleyDeceleratingGrowthTropical2007b; @vlamTemperatureRainfallStrongly2014*].
Potential mechanisms for these patterns include increases in nighttime respiratory carbon loss that disproportionately reduced carbon availability for growth among small trees, or that understory $g_s$  might be stifled under humid conditions, resulting in higher $T_{leaf}$ (Fig. `r fig_leaf_T`) and lower photosynthesis. 
It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures. 
Additional research is required to understand these mechanisms and to disentangle tree growth responses to hot-wet versus hot-dry conditions.
